---
title: "p8130_hw2_ps3194"
author: "Pangsibo Shen"
date: "10/7/2020"
output:
  html_document:
    toc: TRUE
---
## Problem 1

A new test is being developed for the detection of carcinoma of the prostate (Foti et al. N Engl. J
Med. 1977). When it is tested in a group of 113 patients with prostatic cancer, 79 have a positive
diagnosis; in a group of 217 individuals without prostatic cancer, 10 have a positive diagnosis. 

-----------

### 1) Calculate the specificity and sensitivity of the test. 

#### **solution**:

$T$ stands for positive diagnosis of prostatic cancer; $D$ stands for having prostatic cancer
$$ \\ test \ sensitivity = P(T/D) = (79/113) = 0.699 $$
$$ \\ test \ specificity = P(T^{c}/D^{c}) = (207/217) = 0.954 $$

-----------

### 2) In another hypothetical scenario, it is planned to use the test to screen a large sample of subjects for prostatic cancer where the test results will be the only data available. Is this information enough to assess the test characteristics, i.e., sensitivity & specificity? Yes, no, what is missing (if the case)?

#### **solution**:

No, this test will not be enough to provide the test characteristics such as sensitivity and specificity. We need a gold standard test which can tell us which patients in the sample are truly having the disease. 

------------

### 3) In patients with palpable prostatic nodules, the pretest likelihood of prostatic cancer is 50%. The test under these conditions has a sensitivity of 80% and a specificity of 95%. 

#### a) Calculate the probability of a patient with a palpable prostatic nodule and a positive test result having prostatic cancer. What is the epidemiological term of this probability? 

#### **solution**:

By information given, $P(D) = 0.5$ and we want to calculate $P(D/T)$ which has an epidemiological term called Positive Predictive Value(PPV).According to the Bayes'Theorem: 
$$ \\ P(D/T) = \frac{P(T/D)*P(D)}{P(T)} = \frac{P(T/D)*P(D)}{P(T/D)*P(D)+P(T/D^c)*P(D^c)} = \frac{0.8*0.5}{0.8*0.5+0.2*0.5} = 0.8$$

#### b) Re-calculate the probability in 3-a), using a pretest likelihood of prostatic cancer of 10%. Compare the two values and comment on their differences.

#### **solution**:

By information given, $P(D) = 0.1$ and we want to calculate $P(D/T)$ According to the Bayes'Theorem: 
$$ \\ P(D/T) = \frac{P(T/D)*P(D)}{P(T)} = \frac{P(T/D)*P(D)}{P(T/D)*P(D)+P(T/D^c)*P(D^c)} = \frac{0.8*0.1}{0.8*0.1+0.2*0.9} = 0.308$$ In this case, the positive predictive value is significantly smaller than the PPV from 3 a) due to low infection prevalence $P(D)$. In order to have a high positive predictive value, we need to have both high test sensitivity and high infection prevalence. Yet, even if we have a very high test sensitivity, we would still get a poor positive predictive value if the infection prevalence is low.

--------

## Problem 2

According to the Center for Disease Control (CDC), about 34.5% of the adult US population are prediabetic (National Diabetes Statistics Report, 2020). Suppose we randomly select a group of 50 patients seen at Columbia University Medical Center.Calculate the following exact probabilities based on the national statistics:

--------

### 1) Probability that none of these patients are prediabetic.

#### **solution**:

$$P(X=x) = (^{n}_{x})*p^x*(1-p)^{n-x} $$ In this case, we are calculating the probability of X = 0, n = 50 and p = 0.345. 

$$P(X=0) = (^{50}_{0})*0.345^0*(1-0.345)^{50} = 6.48732 *10^{-10} $$

```{r}
dbinom(0, 50, 0.345)
```
Hence the Probability that none of these patients are prediabetic is `r dbinom(0, 50, 0.345)`.

--------

### 2) Probability that less than 10 patients are prediabetic.

#### **solution**:

$$P(X=x) = (^{n}_{x})*p^x*(1-p)^{n-x} $$ In this case, we have X < 10, n = 50 and p = 0.345. 

$$P(X<10) = \sum_{x=0}^9 (^{50}_{x})*0.345^{x}*(1-0.345)^{50-x} =  0.00815 $$

```{r}
pbinom(9,size = 50, prob = 0.345)
```
Hence the Probability that less than 10 patients are prediabetic is `r pbinom(9,size = 50, prob = 0.345)`.

--------

### 3) Probability that 34.5% of these patients are prediabetic (round to the nearest integer). 

#### **solution**:

```{r}
dbinom(18, 50, 0.345)
```
34.5% of 50 patients are 17.25. I rounded it to 18. The Probability that 34.5% of these patients are prediabetic is `r dbinom(17, 50, 0.345)`

-----------

### 4) Could you use an approximation method to calculate the probabilities above? If yes, calculate the probabilities using approximations and compare to the exact values; otherwise, explain why approximations methods are not appropriate. 

#### **solution**:

We can use normal distribution to approximate binomial when $np \geq 10$. In this case, $np = 17.25$ which is greater than 10. We are approximating discrete probability using continuous distribution. A continuity correction should apply. 
$$ P(17.5 \leq X \leq 18.5) $$

```{r}
pnorm(18.5, 17.25, 3.3614) - pnorm(17.5, 17.25, 3.3614)
```

