---
title: "p8130_hw2_ps3194"
author: "Pangsibo Shen"
date: "10/7/2020"
output:
  html_document:
    toc: TRUE
---
## Problem 1

#### A new test is being developed for the detection of carcinoma of the prostate (Foti et al. N Engl. J Med. 1977). When it is tested in a group of 113 patients with prostatic cancer, 79 have a positive diagnosis; in a group of 217 individuals without prostatic cancer, 10 have a positive diagnosis. 

-----------

#### 1) Calculate the specificity and sensitivity of the test. 

#### **solution**:

$T$ stands for positive diagnosis of prostatic cancer; $D$ stands for having prostatic cancer
$$ \\ test \ sensitivity = P(T/D) = (79/113) = 0.699 $$
$$ \\ test \ specificity = P(T^{c}/D^{c}) = (207/217) = 0.954 $$

-----------

#### 2) In another hypothetical scenario, it is planned to use the test to screen a large sample of subjects for prostatic cancer where the test results will be the only data available. Is this information enough to assess the test characteristics, i.e., sensitivity & specificity? Yes, no, what is missing (if the case)?

#### **solution**:

No, this test will not be enough to provide the test characteristics such as sensitivity and specificity. We need a gold standard test which can tell us which patients in the sample are truly having the disease. 

------------

#### 3) In patients with palpable prostatic nodules, the pretest likelihood of prostatic cancer is 50%. The test under these conditions has a sensitivity of 80% and a specificity of 95%. 

#### a) Calculate the probability of a patient with a palpable prostatic nodule and a positive test result having prostatic cancer. What is the epidemiological term of this probability? 

#### **solution**:

By information given, $P(D) = 0.5$ and we want to calculate $P(D/T)$ which has an epidemiological term called Positive Predictive Value(PPV).According to the Bayes'Theorem: 
$$ \\ P(D/T) = \frac{P(T/D)*P(D)}{P(T)} = \frac{P(T/D)*P(D)}{P(T/D)*P(D)+P(T/D^c)*P(D^c)} = \frac{0.8*0.5}{0.8*0.5+0.2*0.5} = 0.8$$

-------------------------------

#### b) Re-calculate the probability in 3-a), using a pretest likelihood of prostatic cancer of 10%. Compare the two values and comment on their differences.

#### **solution**:

By information given, $P(D) = 0.1$ and we want to calculate $P(D/T)$ According to the Bayes'Theorem: 
$$ \\ P(D/T) = \frac{P(T/D)*P(D)}{P(T)} = \frac{P(T/D)*P(D)}{P(T/D)*P(D)+P(T/D^c)*P(D^c)} = \frac{0.8*0.1}{0.8*0.1+0.2*0.9} = 0.308$$ In this case, the positive predictive value is significantly smaller than the PPV from 3 a) due to low infection prevalence $P(D)$. In order to have a high positive predictive value, we need to have both high test sensitivity and high infection prevalence. Yet, even if we have a very high test sensitivity, we would still get a poor positive predictive value if the infection prevalence is low.

--------

## Problem 2

#### According to the Center for Disease Control (CDC), about 34.5% of the adult US population are prediabetic (National Diabetes Statistics Report, 2020). Suppose we randomly select a group of 50 patients seen at Columbia University Medical Center.Calculate the following exact probabilities based on the national statistics:

--------

#### 1) Probability that none of these patients are prediabetic.

#### **solution**:

We can use binomial distribution to calculate the probability for this question.

$$P(X=x) = (^{n}_{x})*p^x*(1-p)^{n-x} $$ In this case, we are calculating the probability of X = 0, n = 50 and p = 0.345. 

$$P(X=0) = (^{50}_{0})*0.345^0*(1-0.345)^{50} = 6.48732 *10^{-10} $$

```{r}
dbinom(0, 50, 0.345)
```
Hence the Probability that none of these patients are prediabetic is `r dbinom(0, 50, 0.345)`.

--------

#### 2) Probability that less than 10 patients are prediabetic.

#### **solution**:

$$P(X=x) = (^{n}_{x})*p^x*(1-p)^{n-x} $$ In this case, we have X < 10, n = 50 and p = 0.345. 

$$P(X<10) = \sum_{x=0}^9 (^{50}_{x})*0.345^{x}*(1-0.345)^{50-x} =  0.00815 $$

```{r}
pbinom(9,size = 50, prob = 0.345)
```
Hence the Probability that less than 10 patients are prediabetic is `r pbinom(9,size = 50, prob = 0.345)`.

--------

#### 3) Probability that 34.5% of these patients are prediabetic (round to the nearest integer). 

#### **solution**:

```{r}
dbinom(18, 50, 0.345)
```
34.5% of 50 patients are 17.25. I rounded it up to 18. The Probability that 34.5% of these patients are prediabetic is `r dbinom(18, 50, 0.345)`

-----------

#### 4) Could you use an approximation method to calculate the probabilities above? If yes, calculate the probabilities using approximations and compare to the exact values; otherwise, explain why approximations methods are not appropriate. 

#### **solution**:

We can use normal distribution to approximate binomial when $np \geq 10$. In this case, $np = 17.25$ which is greater than 10. We are approximating discrete probability using continuous distribution. A continuity correction should apply.The mean of normal distribution will be $n*p = 50*0.345 = 17.25$ and the standard deviation will be $\sqrt{n*p*(1-p)} = \sqrt{50*0.345*0.655} = 3.3614$
$$ P(17.5 \leq X \leq 18.5) $$


```{r}
pnorm(18.5, 17.25, 3.3614) - pnorm(17.5, 17.25, 3.3614)
```

When use normal distribution to approximate binomial, the Probability that 34.5% of these patients are prediabetic is `r pnorm(18.5, 17.25, 3.3614) - pnorm(17.5, 17.25, 3.3614)`

$$ P(0 < X < 9) $$

```{r}
pnorm(9, 17.25, 3.3614) - pnorm(0, 17.25, 3.3614)
```

When use normal distribution to approximate binomial, the Probability that less than 10 patients are prediabetic is `r pnorm(9.5, 17.25, 3.3614) - pnorm(8.5, 17.25, 3.3614)`

$$ P(-0.5 < X < 0.5) $$

```{r}
pnorm(0.5, 17.25, 3.3614) - pnorm(-0.5, 17.25, 3.3614)
```
When use normal distribution to approximate binomial,the Probability that none of these patients are prediabetic is `r pnorm(0.5, 17.25, 3.3614) - pnorm(-0.5, 17.25, 3.3614)`.

------------------

## Problem 3

#### The incidence of uveal melanoma in the US is approximately 5 per million individuals per year, with a significantly higher incidence in non-Hispanic Whites (6.02 per million), when compared to Blacks and Asians: 0.31 and 0.39 per million, respectively.

---------------

#### a) What is the probability that in NYC (population of 8.3 million reported in 2020), exactly 30 cases occur in a given year? 

#### **solution**:

We can use binomial distribution to calculate probability for this question.

$$P(X=30) = (^{8300000}_{30})*(\frac{5}{1000000})^{30}*(1-\frac{5}{1000000})^{8300000-30} = 0.0124 $$

```{r}
dbinom(30, 8300000, 0.000005) #using binomail
dpois(30,41.5)  #using poisson
```
The probability that in NYC, exactly 30 cases occur in a given year is `r dbinom(30, 8300000, 0.000005)`

----------

#### b) Compute the same probability in a) by the mentioned racial/ethnic groups and comment on the findings. Demographic data of NYC in 2020: 14.0% Asians, 42.8% non-Hispanic Whites, 24.3% Black.

#### **solution**:

using the same binomial distribution formula
$$P(X=x) = (^{n}_{x})*p^x*(1-p)^{n-x} $$

```{r}
dbinom(30, 8300000*0.14, (0.39/1000000)) #probability for Asians
dbinom(30, 8300000*0.243, (0.31/1000000)) #probability for Blacks
dbinom(30, 8300000*0.428, (6.02/1000000)) #probability for non-Hispanic Whites
```
The probability that in NYC, exactly 30 cases occur for Asians is `r dbinom(30, 8300000*0.14, (0.39/1000000))`; the probability for Blacks is `r dbinom(30, 8300000*0.243, (0.31/1000000))` ;and The probability for non-Hispanic Whites is `r dbinom(30, 8300000*0.428, (6.02/1000000))`. From the results, we notice that non-Hispanic Whites have the highest probability to have uveal melanoma due to their significantly high incidence rate (6.02 per million) and high proportion living in NYC(42.8%). The other two groups have relatively low probability, because they have relatively low disease incidence rate(0.31 and 0.39 per million) and low proportion living in NYV(24.3% and 14%).

-----------------

## Problem 4

#### People with COVID-19 have had a wide range of symptoms, but high temperature is one of the most important indications. Based on current data, the average temperature of patients diagnosed with COVID-19 follow a normal distribution with a mean of 99.9 degrees Fahrenheit and a standard deviation of 0.73 degrees Fahrenheit. Of course, this differs from person to person, based on factors like body weight, height, the weather, age or gender.Let $𝑋_1, … , 𝑋_{40}$ be the body temperatures of 40 randomly chosen individuals returning to the Columbia University Medical Center. Calculate the following probabilities:


-----------------

#### 1) $P(\bar{𝑋} < 98)$, the probability that the sample mean is less than 98 (average normal temperature).

#### **solution**:

Since the population has a normal distribution, according to the central limit theorem, our sample distribution will have a normal distribution and we can use it to calculate probability. we need to first calculate the standard deviation of the sample means.
$$ \sigma_{\bar{x}}=\frac{\sigma}{\sqrt{n}} $$
$$ \sigma_{\bar{x}}=\frac{0.73}{\sqrt{40}} = 0.115423 $$
$$\mu_{\bar{x}} = \mu = 99.9 $$
```{r}
pnorm(98, mean = 99.9,sd = 0.115423)
```

The probability that the sample mean is less than 98 is `r pnorm(98, mean = 99.9,sd = 0.115423)`

--------

#### 2) $𝑃(\bar{𝑋}> 100.5)$, the probability that the sample mean is greater than 100.5 (alarming zone for COVID-19). 

#### **solution**:

Since the population has a normal distribution, according to the central limit theorem, our sample distribution will have a normal distribution and we can use it to calculate probability. Obtain sample mean and standard deviation from part 1.

```{r}
1 - pnorm(100.5, mean = 99.9,sd = 0.115423)
```

The probability that the sample mean is greater than 100.5 is `r 1 - pnorm(100.5, mean = 99.9,sd = 0.115423)`

-----------------------

#### 3) The 90th quantile of the sampling distribution of the sample mean $\bar{𝑋}$.

#### **solution**:

Since the population has a normal distribution, according to the central limit theorem, our sample distribution will have a normal distribution and we can use it to calculate probability. Obtain sample mean and standard deviation from part 1.

$$ Z = \frac{\bar{X}-\mu}{\sigma}$$

```{r}
qnorm(0.9, mean = 99.9, sd = 0.115423)
z_0.90 = (100.0479 - 99.9)/0.115423
z_0.90
```
The 90th quantile of the sampling distribution of the sample mean $\bar{𝑋}$ is `r qnorm(0.9, mean = 99.9, sd = 0.115423)` with a Z score of `r z_0.90`

---------------------------

#### 4) The cutoff values for the middle 50% of the sampling distribution of the sample mean $\bar{𝑋}$.

#### **solution**:

Since the population has a normal distribution, according to the central limit theorem, our sample distribution will have a normal distribution and we can use it to calculate probability. Obtain sample mean and standard deviation from part 1.

$$ Z = \frac{\bar{X}-\mu}{\sigma}$$
$$ Z_{0.75} = \frac{99.97785-99.9}{0.115423} = 0.674$$
$$ Z_{0.25} = \frac{99.82215-99.9}{0.115423} = -0.674$$

```{r}
qnorm(0.75, mean = 99.9, sd = 0.115423)
z_0.75 = (99.97785-99.9)/0.115423
z_0.75
qnorm(0.25, mean = 99.9, sd = 0.115423)
z_0.25 = (99.82215-99.9)/0.115423
z_0.25
```
The cutoff values for the middle 50% of the sampling distribution of the sample mean $\bar{𝑋}$ are  Z_0.75 (`r z_0.75`, x = 99.978) and Z_0.25 (`r z_0.25`, x = 99.822).
